\textbf{ Interpretable Gaussian Processes for Stellar Variability\\ I: Static Surfaces }

### 2020-12-29 tex/ms.tex
adding tutorial
docs fix; more text

We place uniform priors on all five quantities, with support in
$1 \leq n \leq 50$, $0 < c \leq 1$,
$10^\circ \leq r \leq 30^\circ$,
$0 \leq a \leq 1$, and
$0 \leq b \leq 1$. Note that while $n$ formally represents an integer,
its effect on the GP is a scaling of the covariance (see Equation~\ref{eq:exp_yy_sep});
as such, it has support over all real numbers within the bounds listed
above. We \emph{could} restrict it to integer values, but this would make
sampling quite tricky. Moreover, in practice it is useful to allow for
noninteger values to add some flexibility to the model; we discuss this
in more detail in \S\ref{sec:integerspots}

@@

The reason for this bias has to do with the baseline degeneracy discussed in
\S\ref{sec:baseline} and illustrated in Figure~\ref{fig:mean_normalization}:
the total spottiness of a stellar surface cannot be probed in relative
photometry. In particular, this means spots near the poles lie almost
entirely in the null space. Applied to the problem at hand, this degeneracy
makes it difficult to distinguish between stars with spots concentrated
exclusively at mid-latitudes (in this case, the truth) and
stars with spots centered closer to the poles but with large latitudinal
variance. The latter configuration leads to many spots close to the poles,
whose effect on the (relative) light curve is negligible, and some spots
at mid-latitudes, whose effect on the light curve is similar to that of
the former configuration.
Thus, the data alone cannot be used to discriminate between these two
scenarios, introducing the degeneracy we see in the posterior. In fact,
it is clear that in the tails of the distribution, the mean spot latitude
and the standard deviation of spot latitudes are positively correlated.
The bias we see is therefore not a shortcoming of the model, but of the
\emph{data} itself. To get around this, we either need to impose a
shrinkage prior on $\sigma_\phi$ (\S\ref{sec:sigma-prior}),
observe in a different mode that is sensitive to baseline information
(\S\ref{sec:breaking_baseline_degeneracy}), or simply collect more data.
As we will see below, the degeneracy described above is not perfect:
for very large ($M \sim 1{,}000$) ensembles, high-variance polar spots
can confidently be ruled out.

@@

\subsection{Small spots}
\label{sec:tinyspots}

\xxx{TODO: Discuss what we should do about tiny spots.}

\subsection{The baseline degeneracy}
\label{sec:breaking_baseline_degeneracy}

\xxx{TODO: Discuss the baseline degeneracy and how to break it. Discuss the
    degeneracy between $c$ and $N$ and how to break it.}

\subsection{Gaussians all the way down}
\label{sec:other-marg}
\xxx{TODO: We can also marginalize over period and limb darkening coefficients.}

\subsection{Caveats}
\label{sec:caveats}
\xxx{
    \begin{enumerate}
        \item The latitude distribution is not Gaussian.
        \item The number of spots does not have to be an integer.
        \item Careful with light curves with large amplitudes, although
              Figure~\ref{fig:calibration_hicontrast}
              has max amplitudes of more than 10\% and works great.
        \item No differential rotation / spot evolution yet!
    \end{enumerate}
}

\subsection{Other distributions}
\label{sec:other-dist}
\xxx{TODO: Discuss the uniform spot radius distribution.}

\subsection{Prior on sigma}
\label{sec:sigma-prior}
\xxx{TODO: We might want a shrinkage prior on sigma.}

\subsection{Inferring inclinations}
\xxx{TODO: Works great. Note that inclination distribution is sometimes bimodal!}

@@

We implement our interpretable GP in the open-source, user-friendly
\Python package \starryprocess, which can be installed via \textsf{pip}\xspace
or from source on GitHub\footnote{\url{https://github.com/rodluger/starry_process}}.
The code is thoroughly unit-tested and well documented, with examples on how
to use the GP in custom inference problems. Users can choose, among other
options, whether or not
to marginalize over inclination (\S\ref{sec:inclination}) and
whether or not to model a normalized
process (\S\ref{sec:baseline}). Users can also choose the spherical harmonic
degree of the expansion, although it is recommended to use $l_\mathrm{max} = 15$
(see below).
Users may compute the mean vector and covariance matrix in either the spherical
harmonic basis or the flux basis, or they may sample from it or use it to
compute marginal likelihoods.
Arbitrary order limb darkening is implemented following
\citet{Agol2020}.

@@

In ensemble analyses, we must compute the likelihood of each of the $M$ light curves
conditioned on $\pmb{\theta}_\bullet$. In practice, each star will have a
different rotation period, different limb darkening coefficients, and different
photometric uncertainty, meaning we must factorize $M$ different covariance
matrices. Fortunately, the spherical harmonic covariance, $\pmb{\Sigma}_\mathbf{y}$
(\S\ref{sec:gp}) need only be computed once; this can then be linearly
transformed into the flux basis for each light curve. As the evaluation of
$\pmb{\Sigma}_\mathbf{y}$ is the computationally-intensive step, the
likelihood evaluation typically scales sub-linearly with $M$. Furthermore,
as we discussed in \ref{sec:other-marg}, it is possible to marginalize over
the period and limb darkening coefficients, which would remove the scaling
with $M$ entirely if the photometric precision were the same for all
light curves. Even if it is not, the algorithm can still be greatly sped up,
although an implementation of this is deferred to the next paper.
\xxx{DFM: Is this true? I.e., is there an efficient way to solve Equation~(\ref{eq:log-like})
    when $\pmb{\Sigma}$ is the same for all stars but $\mathbf{C}$ is different? Is there
    some limit in which an approximation of some sort can be made?}

@@

\section{Conclusions}
\label{sec:conclusions}
\xxx{TODO: Summarize. Discuss future work, including differential rotation
    and spot evolution.}

>>>>

We place uniform priors on all five quantities, with support in
$1 \leq n \leq 50$, $0 < c \leq 1$,
$10^\circ \leq r \leq 30^\circ$,
$0 \leq a \leq 1$, and
$0 \leq b \leq 1$. Note that while $n$ formally represents an integer,
its effect on the GP is a scaling of the covariance (see Equation~\ref{eq:exp_yy_sep});
as such, it has support over all real numbers within the bounds listed
above. We \emph{could} restrict it to integer values, but this would make
sampling quite tricky. Moreover, in practice it is useful to allow for
noninteger values to add some flexibility to the model; we discuss this
in more detail in \S\ref{sec:caveats}

@@

The reason for this bias has to do with the baseline degeneracy discussed in
\S\ref{sec:baseline} and illustrated in Figure~\ref{fig:mean_normalization}:
the total spottiness of a stellar surface cannot be probed in relative
photometry. In particular, this means spots near the poles lie almost
entirely in the null space. Applied to the problem at hand, this degeneracy
makes it difficult to distinguish between stars with spots concentrated
exclusively at mid-latitudes (in this case, the truth) and
stars with spots centered closer to the poles but with large latitudinal
variance. The latter configuration leads to many spots close to the poles,
whose effect on the (relative) light curve is negligible, and some spots
at mid-latitudes, whose effect on the light curve is similar to that of
the former configuration.
Thus, the data alone cannot be used to discriminate between these two
scenarios, introducing the degeneracy we see in the posterior. In fact,
it is clear that in the tails of the distribution, the mean spot latitude
and the standard deviation of spot latitudes are positively correlated.
The bias we see is therefore not a shortcoming of the model, but of the
\emph{data} itself. To get around this, we either need to impose a
shrinkage prior on $\sigma_\phi$ (\S\ref{sec:caveats}),
observe in a different mode that is sensitive to baseline information
(\S\ref{sec:breaking_baseline_degeneracy}), or simply collect more data.
As we will see below, the degeneracy described above is not perfect:
for very large ($M \sim 1{,}000$) ensembles, high-variance polar spots
can confidently be ruled out.

@@

\subsection{Small spots}
\label{sec:tinyspots}
\xxx{TODO: The model is limited in resolution to features with radii $r \gtrsim 10^\circ$.
    Typical sun spots are MUCH smaller than that. However, our model can still
    correctly infer latitudinal parameters even when spots are really tiny
    (i.e., no bias due to the limited resolution). We might try to improve the
    numerical stability of the model at higher spherical harmonic degree, but
    that's a task for a future paper.}

\subsection{Bright spots}
\label{sec:brightspots}
\xxx{TODO: We can easily model bright spots and plages by setting $c < 0$. Show
    an example. This model can be used to differentiate between spot-dominated
    and plage-dominated surfaces, which is physically interesting. Cite
    Brett's paper on TRAPPIST-1.}

\subsection{The baseline degeneracy}
\label{sec:breaking_baseline_degeneracy}
\xxx{TODO: Discuss the baseline degeneracy and how to break it. Discuss the
    degeneracy between $c$ and $N$ and how to break it. The basic idea here
    is that if we can constrain the (spotless) photospheric temperature, we
    can uniquely infer $n$. I think this can be done by taking light curves
    in two bands, but I'm checking with Gully. @HEDGES: Anything to add on this?}

\subsection{Gaussians all the way down}
\label{sec:other-marg}
\xxx{TODO: We can also marginalize over period and limb darkening coefficients.
    Argue that it's just integrals of Wigner matrices all over again, which we
    can compute efficiently in closed form. I think we need a uniform prior on
    the rotation \emph{frequency}, which is probably fine. And I think we can
    parametrize limb darkening in terms of Kipping's $q_1$ and $q_2$ and place
    a uniform prior in that space---I'm fairly confident the integrals can still
    be solved exactly.}

\subsection{Time evolution}
\label{sec:temporal}
\xxx{TODO: The model presented in this paper assumes statics stellar surfaces,
    corresponding to perfectly periodic light curves. This is hardly ever the
    case in practice, as spots evolve and differential rotation changes the surface
    over time. We can still use this model, but we need to either restrict our analysis
    to segments of light curves of duration shorter than the timescale on which
    the surface evolves, or explicitly model the effect of time evolution on the covariance.
    The next paper will deal with temporal evolution more
    rigorously, but the way to do this is just take the elementwise product of the \starryprocess
    flux covariance with a temporal covariance. This corresponds to the Kronecker product
    in spherical harmonic space. Show a figure, give an example, etc.}

\subsection{Application to transits and RV}
\label{sec:other-work}
\xxx{TODO: The covariance in the spherical harmonic basis can be used as a prior
    in many other settings.}

\subsection{Comparison to other work}
\label{sec:other-work}
\xxx{TODO: Discuss \citet{Basri2020}, how we agree and disagree. Compare to Brett's
    paper using ABC to measure spot sizes as a function of spectral type. Discuss
    Simon Wood's paper in ecology on synthetic likelihoods. Talk about similarities
    to Gaussian Random Fields in CMB studies.}

\subsection{Caveats}
\label{sec:caveats}
\xxx{TODO: Miscellaneous caveats.}
\xxx{
    \begin{enumerate}
        \item The latitude distribution is not Gaussian, so careful
              when interpreting $\mu_\phi$ and $\sigma_\phi$
        \item The number of spots does not have to be an integer. This is
              fine, since it makes the model more flexible.
        \item Careful with light curves with large amplitudes, although
              Figure~\ref{fig:calibration_hicontrast}
              has max amplitudes of more than 10\% and works great.
        \item No differential rotation / spot evolution yet! We're working on it.
        \item We might want a shrinkage prior on sigma, since it's hard to
              break the polar spot degeneracy with single-band light curves.
    \end{enumerate}
}

@@

We implement our interpretable GP in the open-source, user-friendly
\Python package \starryprocess, which can be installed via \textsf{pip}\xspace
or from source on \href{https://github.com/rodluger/starry_process}{GitHub}.
The code is thoroughly unit-tested and well documented, with examples on how
to use the GP in custom inference problems. Users can choose, among other
options, whether or not
to marginalize over inclination (\S\ref{sec:inclination}) and
whether or not to model a normalized
process (\S\ref{sec:baseline}). Users can also choose the spherical harmonic
degree of the expansion, although it is recommended to use $l_\mathrm{max} = 15$
(see below).
Users may compute the mean vector and covariance matrix in either the spherical
harmonic basis or the flux basis, or they may sample from it or use it to
compute marginal likelihoods.
Arbitrary order limb darkening is implemented following
\citet{Agol2020}.

@@

In ensemble analyses, we must compute the likelihood of each of the $M$ light curves
conditioned on $\pmb{\theta}_\bullet$. In practice, each star will have a
different rotation period, different limb darkening coefficients, and different
photometric uncertainty, meaning we must factorize $M$ different covariance
matrices. Fortunately, the spherical harmonic covariance, $\pmb{\Sigma}_\mathbf{y}$
(\S\ref{sec:gp}) need only be computed once; this can then be linearly
transformed into the flux basis for each light curve. As the evaluation of
$\pmb{\Sigma}_\mathbf{y}$ is the computationally-intensive step, the
likelihood evaluation typically scales sub-linearly with $M$. Furthermore,
as we discussed in \ref{sec:other-marg}, it is possible to marginalize over
the period and limb darkening coefficients, which would remove the scaling
with $M$ entirely if the photometric precision were the same for all
light curves. Even if it is not, the algorithm can still be greatly sped up,
although an implementation of this is deferred to the next paper.
\xxx{@DFM: Is this true? I.e., is there an efficient way to solve Equation~(\ref{eq:log-like})
    when $\pmb{\Sigma}$ is the same for all stars but $\mathbf{C}$ is different? Is there
    some limit in which an approximation of some sort can be made?}

@@

\section{Conclusions}
\label{sec:conclusions}
We presented a new Gaussian process (GP) model for stellar variability
whose hyperparameters explicitly correspond to physical properties of the stellar
surface. Our GP allows one to efficiently compute the likelihood function
for stellar light curves
marginalized over nuisance parameters such as the specific sizes,
positions, and contrasts of individual spots,
which are generally unknowable due to the extreme degeneracies
involved in the light curve mapping problem. Our GP therefore
makes it easy to
do posterior inference on the real quantities of interest: parameters
controlling the \emph{distribution} of spot sizes, latitudes, and
contrasts within a star and/or across many stars in an ensemble.

Because our expression for the GP covariance has an exact, closed-form
solution as a function of the spot parameters, it can be computed
extremely efficiently: a typical likelihood evaluation on a dataset
consisting of $K \sim 1{,}000$ points takes about 10ms on a modern
laptop. Our algorithm is implemented in the open-source, user-friendly
\Python package \starryprocess, which is \textsf{pip}-installable and available on
\href{https://github.com/rodluger/starry_process}{GitHub}.
In keeping with other papers in the \starry series, all figures in this
paper are generated automatically from open-source scripts linked to in
each of the captions \codeicon, and the principal equations link to associated
unit tests that ensure the accuracy and reproducibility of the algorithm
presented here \testpassicon/\testfailicon.

We devoted a large portion of this paper to testing the algorithm on a variety
of synthetic datasets, showing that it is a well-calibrated and in most
cases unbiased estimator for star spot properties. Below we list the main
results presented in this paper:

\begin{enumerate}
    \item \textbf{Our GP works best for ensemble analyses.} The light curve mapping
          problem is extremely degenerate, as light curves contain a vanishingly small
          fraction of the total information about a stellar surface. However, the
          degenerate surface modes are a strong function of the observer's viewing
          angle, so light curves of stars seen at different inclinations constrain
          different components of the surface. We have shown that if we jointly
          analyze the light curves of many stars,
          we can break many of the degeneracies at play and uniquely infer the
          statistical properties of the spots across the ensemble. This type
          of analysis works best if the stars in the ensemble are statistically
          similar: i.e., the properties of their spots are all drawn from the
          same parent distribution, whose parameters we can constrain.
    \item \textbf{In general, an ensemble of at least $M \sim 50$ light curves
              is needed to place meaningful constraints on the spot properties}.
          This estimate is based on ensemble analyses of light curves with
          $K = 1{,}000$ cadence each and per-cadence precision of one part
          per thousand. Lower quality, shorter baseline observations, or
          datasets contaminated by outliers will in general require larger values
          of $M$ for the same constraining power. The presence of strong limb
          darkening also degrades the information content of light curves, in
          which case an ensemble of hundreds or even a thousand light curves
          is recommended.
    \item \textbf{Our GP is an unbiased estimator for the spot radius and
              latitude distributions.} We showed that our GP can accurately infer the
          angular size of spots and the mode and standard deviation of their
          distribution in latitude from stellar light curves. For the fiducial
          ensemble of $M = 50$ light curves described above, we are able to
          constrain the average spot radii to within a couple degrees and the
          average spot latitudes to within $5^\circ$.
    \item \textbf{Relative single-band photometry cannot uniquely constrain
              the spot contrast or the total number of spots}. These two quantities
          are degnerate with each other. While the contrast can be very weakly
          constrained with lots of data, the total number of spots is not
          an observable in relative single-band photometry. Knowledge of the
          unspotted photospheric brightness (or temperate) can break this
          degeneracy, so analyses of multi-band light curves can in principle
          infer the total spottiness of a stellar surface.
    \item \textbf{Our GP can accurately infer stellar inclinations.} We presented
          two versions of our GP: one conditioned on a specific value of the stellar
          inclination, and one marginalized over inclination under an isotropic
          prior. In both cases, we find that we can accurately infer the inclinations
          of individual stars in an ensemble analysis (in the latter case, a
          simple post-processing step can yield the inclination posterior distribution).
          While the inclination is not an observable for an individual stellar light
          curve, the population-level constraints on the spot properties achieved
          by the GP can break the degeneracies involving the inclination, allowing
          us to usually infer it to within about $10^\circ$ and without bias.
    \item \textbf{Our GP can be used to model small, Sun-like spots.} The algorithm
          presented here is limited to a surface resolution of about $10^\circ$,
          corresponding to spots about an order of magnitude larger than typical
          Sun spots. However, we have showed that when we apply our model to
          light curves of stars with small ($r \sim 3^\circ$) spots, we can
          still infer their latitudinal distribution without bias.
    \item \textbf{Our GP can be extended to model time-variable surfaces.} The
          algorithm presented here was derived for static stellar surfaces,
          corresponding to perfectly periodic light curves. \xxx{...}
    \item \textbf{Possible applications.} Spot-dominated versus plage-dominated
\end{enumerate}

\xxx{TODO: Discuss future work, including differential rotation
    and spot evolution.}
